XXX in a Tracing JIT Compiler for Efficient Dynamic Languages

### 2011-03-24 talk/icooolps2011/paper.tex
more figureyfication

Here's a diagram of this process:

@@

On the left you see the levels of execution. The CPU executes the binary of
PyPy's Python interpreter, which consists of RPython functions that have been
compiled first to C, then to machine code. Some of these functions contain
loops, others don't. The interpreter runs a Python program written by a
programmer (the user). If the tracer is used, it traces operations on the level
of the interpreter. However, the extent of the trace is determined by the loops
in the user program.

@@

Let's assume we trace through code that sums three attributes, such as:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{n}{inst}\PY{o}{.}\PY{n}{getattr}\PY{p}{(}\PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)} \PY{o}{+} \PY{n}{inst}\PY{o}{.}\PY{n}{getattr}\PY{p}{(}\PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)} \PY{o}{+} \PY{n}{inst}\PY{o}{.}\PY{n}{getattr}\PY{p}{(}\PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\end{Verbatim}

The trace could look like this:

@@

In this example, the attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The trace indeed contains
five calls to \texttt{dict.get}, which is slow.

@@

With this changed instance implementation, the trace we had above changes to the
following, where \texttt{0xb74af4a8} is the memory address of the Map instance that
has been promoted:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{c}{# inst.getattr("a")}
\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index1} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index1} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{n}{index1}\PY{p}{]}

\PY{c}{# inst.getattr("b")}
\PY{n}{map2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map2} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index2} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index2} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods1} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
\PY{n}{result2} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}

\PY{c}{# inst.getattr("c")}
\PY{n}{map3} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map3} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index3} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map3}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index3} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods2} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
\PY{n}{result3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}

\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{n}{result3}
\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
\end{Verbatim}

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
a pure function and they have constant arguments. That means that \texttt{index1/2/3}
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. The optimized trace looks like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{c}{# inst.getattr("a")}
\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}

\PY{c}{# inst.getattr("b")}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods1} \PY{o}{=} \PY{n}{cls1}\PY{o}{.}\PY{n}{methods}
\PY{n}{result2} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}

\PY{c}{# inst.getattr("c")}
\PY{n}{cls2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods2} \PY{o}{=} \PY{n}{cls2}\PY{o}{.}\PY{n}{methods}
\PY{n}{result3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}

\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{n}{result3}
\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
\end{Verbatim}

@@

Therefore, we give every class a version number, which is increased every time a
class gets changed (i.e., the content of the \texttt{methods} dictionary changes).
This means that the result of \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a pure operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} looks like
this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{VersionTag}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{pass}

\PY{k}{class} \PY{n+nc}{Class}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{name} \PY{o}{=} \PY{n}{name}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version} \PY{o}{=} \PY{n}{VersionTag}\PY{p}{(}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n}{version} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{n}{version}\PY{p}{)}
        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
            \PY{k}{return} \PY{n}{result}
        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{version}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{change\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version} \PY{o}{=} \PY{n}{VersionTag}\PY{p}{(}\PY{p}{)}
\end{Verbatim}

@@

The trace with this new class implementation looks like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{c}{# inst.getattr("a")}
\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index1} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index1} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{n}{index1}\PY{p}{]}

\PY{c}{# inst.getattr("b")}
\PY{n}{map2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map2} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index2} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index2} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{guard}\PY{p}{(}\PY{n}{cls1} \PY{o}{==} \PY{l+m+mh}{0xb7aaaaf8}\PY{p}{)}
\PY{n}{version1} \PY{o}{=} \PY{n}{cls1}\PY{o}{.}\PY{n}{version}
\PY{n}{guard}\PY{p}{(}\PY{n}{version1} \PY{o}{==} \PY{l+m+mh}{0xb7bbbb18}\PY{p}{)}
\PY{n}{result2} \PY{o}{=} \PY{n}{Class}\PY{o}{.}\PY{n}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n}{cls}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{,} \PY{n}{version1}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}

\PY{c}{# inst.getattr("c")}
\PY{n}{map3} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map3} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index3} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map3}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index3} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{cls2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{guard}\PY{p}{(}\PY{n}{cls2} \PY{o}{==} \PY{l+m+mh}{0xb7aaaaf8}\PY{p}{)}
\PY{n}{version2} \PY{o}{=} \PY{n}{cls2}\PY{o}{.}\PY{n}{version}
\PY{n}{guard}\PY{p}{(}\PY{n}{version2} \PY{o}{==} \PY{l+m+mh}{0xb7bbbb18}\PY{p}{)}
\PY{n}{result3} \PY{o}{=} \PY{n}{Class}\PY{o}{.}\PY{n}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n}{cls}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{,} \PY{n}{version2}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}

\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{n}{result3}
\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
\end{Verbatim}

The calls to \texttt{Class.\_find\_method} can now be optimized away, also the
promotion of the class and the version, except for the first one. The final
optimized trace looks like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{c}{# inst.getattr("a")}
\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}

\PY{c}{# inst.getattr("b")}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{guard}\PY{p}{(}\PY{n}{cls1} \PY{o}{==} \PY{l+m+mh}{0xb7aaaaf8}\PY{p}{)}
\PY{n}{version1} \PY{o}{=} \PY{n}{cls1}\PY{o}{.}\PY{n}{version}
\PY{n}{guard}\PY{p}{(}\PY{n}{version1} \PY{o}{==} \PY{l+m+mh}{0xb7bbbb18}\PY{p}{)}
\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{l+m+mi}{41}

\PY{c}{# inst.getattr("c")}
\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{l+m+mi}{17}
\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
\end{Verbatim}

>>>>

@@

Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left you
see the levels of execution. The CPU executes the binary of
PyPy's Python interpreter, which consists of RPython functions that have been
compiled first to C, then to machine code. Some of these functions contain
loops, others don't. The interpreter runs a Python program written by a
programmer (the user). If the tracer is used, it traces operations on the level
of the interpreter. However, the extent of the trace is determined by the loops
in the user program.

@@

Let's assume we trace through code that sums three attributes, such as:

\begin{Verbatim}
inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
\end{Verbatim}

@@

The trace would look like in Figure~\ref{fig:trace1}. In this example, the
attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The trace indeed contains
five calls to \texttt{dict.get}, which is slow.

@@

With this changed instance implementation, the trace we had above changes to the
following, where \texttt{0xb74af4a8} is the memory address of the Map instance that
has been promoted, see Figure~\ref{fig:trace2}.

\begin{figure}
\input{code/trace2.tex}
\caption{Unoptimized Trace After the Introduction of Maps}
\label{fig:trace2}
\end{figure}

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
a pure function and they have constant arguments. That means that \texttt{index1/2/3}
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. The optimized trace looks can be seen in Figure~\ref{fig:trace3}

\begin{figure}
\input{code/trace3.tex}
\caption{Optimized Trace After the Introduction of Maps}
\label{fig:trace3}
\end{figure}

@@

Therefore, we give every class a version number, which is increased every time a
class gets changed (i.e., the content of the \texttt{methods} dictionary changes).
This means that the result of \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a pure operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} looks like
in Figure~\ref{fig:version}

\begin{figure}
\input{code/version.tex}
\caption{Versioning of Classes}
\label{fig:version}
\end{figure}

@@

\begin{figure}
\input{code/trace4.tex}
\caption{Unoptimized Trace After Introduction of Versioned Classes}
\label{fig:trace4}
\end{figure}

The trace with this new class implementation can be seen in
Figure~\ref{fig:trace4}.
The calls to \texttt{Class.\_find\_method} can now be optimized away, also the
promotion of the class and the version, except for the first one. The final
optimized trace can be seen in Figure~\ref{fig:trace5}.

\begin{figure}
\input{code/trace5.tex}
\caption{Optimized Trace After Introduction of Versioned Classes}
\label{fig:trace5}
\end{figure}
